Staged Abstract Interpreters

### 2019-02-22 paper/10_intro.tex
paper

If the programmer identifies that @b@ will be known in the future stage (as shown
on its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known in the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@ and generate a specialized function that runs faster:

\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b * 1
\end{lstlisting}

Now consider that we have an interpreter @eval@ for some language as following:

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from \note{monadic interpreters}
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979} and embedded domain-specific languages
\cite{DBLP:conf/snapl/RompfBLSJAOSKDK15, DBLP:journals/jfp/CaretteKS09,
DBLP:conf/icfp/GibbonsW14}, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework. The four semantics
share one same generic interpreter. However, by instantiating the interpreter
differently, it can behave as a concrete interpreter, an abstracter interperter,
or their staged versions, respectively. The key idea to enable this flexibility
is to use abstract type members \note{or type parameters} abstract over concrete/abstract components
(e.g., concrete values or abstract values) of the interpreter, as well as the
binding-time (e.g., static or dyanmic).

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static
analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and \textit{efficiency}, we argue
that applying meta-programming (multi-stage programming, in particular) with
abstract interpreters is an effective approach to improve the latter one while
the former one is untouched. Specialization removes the overhead of abstractions
introduced by high-level programming, but not changes the meaning of the program
-- in this case, the abstract semantics implemented by the analyzer. Indeed,
specializing analysis with respect to a program by partial evaluation
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new idea, despite the fact that most of the previous
work are ad-hoc on a particular analysis or requiring significant changes to the
analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

>>>>

If the programmer identifies that @b@ will be known in the future stage (as shown
on its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known at the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@, and generate a specialized function where the overhead of
recursion is eliminated. The following is the generated code:

\begin{lstlisting}
def power5(b: Int) = {
  val x1 = x0 * x0; val x2 = x0 * x1; val x3 = x0 * x2; val x4 = x0 * x3
  x4
}
\end{lstlisting}

Now to concretely illustrate the idea of Futamura projection, consider that we
have an interpreter @eval@ of some language as following:

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14}, we develop an
approach to fulfill four different semantics (Figure~\ref{confluence}) in a
unified framework. The four semantics share one same generic interpreter.
However, by instantiating the interpreter differently, it can behave as a
concrete interpreter, an abstracter interperter, or their code-generator
versions, respectively. The key idea that enables this flexibility is to use
abstract type members abstract over concrete/abstract components (e.g., concrete
values or abstract values), as well as the binding-time of them (e.g., static or
dyanmic). By staging, the overhead caused by the monadic layers is eliminated in
the generated code.

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textbf{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming, but
not changes the meaning of the program -- in this case, the abstract semantics
implemented by the analyzer. Indeed, specializing analysis with respect to a
program by partial evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new idea, despite the
fact that most of the previous work are ad-hoc on a particular analysis or
requiring significant changes to the analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

### 2019-02-22 paper/70_related.tex
misc
paper

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The big-step abstract interpreters we presented in this
paper are also inspired by the AAM framework, which are later adopted for
big-step abstract interpreters \cite{DBLP:journals/pacmpl/DaraisLNH17,
Wei:2018:RAA:3243631.3236800}. \cite{DBLP:conf/cc/CousotC02} also proposed an
abstract interpretation framework for modular analysis.
\citet{DBLP:conf/popl/CalcagnoDOY09} developed compositional shape analysis by
using bi-abduction. The big-step abstract interpreter we presented in this paper
is inspired by \cite{DBLP:journals/pacmpl/DaraisLNH17,
Wei:2018:RAA:3243631.3236800}. \citet{Keidel:2018:CSP:3243631.3236767}
demonstrate a similar implementation of definitional abstract interpreter using
arrows.

@@

\cite{10.1007/11561347_18}

>>>>

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17}.
Similar to monadic abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present an arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\cite{DBLP:conf/cc/CousotC02} proposed an
abstract interpretation framework for modular analysis.
\citet{DBLP:conf/popl/CalcagnoDOY09} developed compositional shape analysis by
using bi-abduction.

@@

MSP with functors and monads \cite{10.1007/11561347_18}

staged monadic interpreters \cite{DBLP:conf/dsl/SheardBP99}
